### 2015-02-23 doc/thesis/evaluation.tex

LuaSocket \citep{luasocket} and LuaFileSystem \citep{luafilesystem} were
the third and fourth case studies that we used to evaluate Typed Lua.
We chose them because they are the most popular Lua libraries.
We wrote a script that builds the dependency graph of Lua libraries
that are in the LuaRocks repository, and used this dependency graph
to identify the most popular Lua libraries.

We also randomly selected three case studies from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first provides client side HTTP digest authentication for Lua.
The second is an extension to the primitive function \texttt{type}.
The third is a generator and checker of modulo 11 numbers.
We randomly selected three case studies because we wanted to evaluate
Typed Lua for annotating existing libraries that are writting in Lua,
as the previous case studies are mostly libraries that are writting in C.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
Lua Standard Library & 62\% & 5\% & 6\% & 27\% & 129 \\ % 100%
\hline
MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
\hline
LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them include,
and we discuss the contributions and limitations that these
evaluation results represent in our type system.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
base & 34\% & 4\% & 8\% & 54\% & 26 \\ % 100%
\hline
coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
\hline
package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
\hline
string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
\hline
utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
\hline
table & 15\% & 71\% & 0\% & 14\% & 7 \\ % 100%
\hline
math & 81\% & 0\% & 15\% & 4\% & 27 \\ % 100%
\hline
io & 71\% & 0\% & 0\% & 29\% & 21 \\ % 100%
\hline
os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

We could give precise static types to several members of
the Lua Standard Library, though some of them are still
hard to type with our current type system.
Now we will discuss why some modules are difficult to type,
and what kind of limitations they represent in our type system.

The \texttt{base} module was quite hard to type because it
includes several functions that rely on reflection.
For instance, the function \texttt{pairs} traverses all
keys and values that are stored in a given table.
As another example, the function \texttt{getmetatable}
returns the metatable of a given table.

We could not give a precise static type to \texttt{ipairs},
because our type system does not have parametric polymorphism.
This function returns an iterator function that traverses a
list of elements.

The \texttt{base} module also includes two overloaded functions:
\texttt{tonumber} and \texttt{collectgarbage}.

@@

We could not type most of the members of the \texttt{coroutine} module
because our type system cannot describe the computational effects of a program.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.
Still, we could type the function \texttt{coroutine.isyieldable},
as it has no input parameters, and simply returns a boolean that
indicates whether the running coroutine is yieldable.

We could type most of the members of the \texttt{package} module,
but we could not type the following tables: \texttt{package.loaded},
\texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types rely on reflection,
that is, their types depend on the modules a program loads.

The \texttt{string} module was not that hard to type, but it includes
some functions that we could not type, as they rely on format
strings.
For instance, the type of the arguments that we pass to
\texttt{string.format} depend on the format string we are using.
It is fine to call \texttt{string.format("\%d", 1)}, but
\texttt{string.format("\%d", true)} raises a run-time error.

Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style, and we treat this feature as a
special case in the implementation.
For this reason, we did not include the typing of the \texttt{string}
methods in our evaluation results.

@@

The \texttt{table} module was specially difficult to type because
most of its functions require parametric polymorphism.
These functions either receive or return a list of elements, and
parametric polymorphism would help us to describe them with a generic type.

However, the lack of parametric polymorphism did not prevent us from
giving a precise type to \texttt{table.concat}, as it operates over lists
where all elements are strings or numbers.

Even if our type system had parametric polymorphism, it would
be difficult to type \texttt{table.insert}, as its type depends on
the calling arity.
More precisely, calling \texttt{table.insert(l, v)} inserts the
element \texttt{v} at the end of the list \texttt{l},
while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
at the position \texttt{p} of the list \texttt{l},
but generates a run-time error when \texttt{p} is out of bounds.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

Even though the \texttt{math} module looks straightforward to type,
it includes several overloaded functions.
For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.

The \texttt{math} module also includes a function that is hard to type:
the function \texttt{math.random}.
It is difficult to type because its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
This function also does not follow the semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

The \texttt{io} module provides operations for manipulating files,
and these operations can use implicit or explicit file descriptors.
The implicit operations are functions in the \texttt{io} table,
while the explicit operations are methods of a file descriptor.
We used an \emph{userdata} declaration to introduce the type
\texttt{file} for representing the type of a file descriptor
and its methods.
For this reason, we included the typing of the \texttt{file}
methods in our evaluation results.

We could type most of members of the \texttt{io} module,
but we could not precisely type the functions \texttt{io.close},
\texttt{io.read}, and \texttt{io.lines}.
We also could not type the methods \texttt{file:close},
\texttt{file:read}, and \texttt{file:lines}.

The function \texttt{io.close} is difficult to type
because its return type depends on whether the file handle that is
being closed was created with \texttt{popen}.
Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
when \texttt{f} was created with \texttt{popen},
and it returns the tuple \texttt{(boolean)?} otherwise.
The method \texttt{file:close} is difficult to type for the same reason.

The functions \texttt{io.read} and \texttt{io.lines} are difficult
to type because they are overloaded functions.
For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
The function \texttt{io.lines}, and the methods \texttt{file:read} and
\texttt{file:lines} are difficult to type for the same reason.

The function \texttt{os.execute} is the only member of the \texttt{os}
module that we could not give a precise type, as it is an overloaded
function.
More precisely, \texttt{os.execute()} returns a boolean, while
\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
where \texttt{c} is a string that represents the command to be executed.

@@

We could type most of the members in the \texttt{socket} module,
which is the C core.
However, this module includes some functions such as
\texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
that are hard to type because they rely on reflection.
For instance, we can use \texttt{socket.skip} to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} does not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \citep{flanagan2006htc}.

We could type most of the members of the modules \texttt{ftp} and
\texttt{http}, but we could not precisely type
the functions \texttt{ftp.get} and \texttt{http.request} because
they are overloaded functions.

@@

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encoding.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low level
filters to create standard filters.

@@

Even though we could type most of the functions exported by the
\texttt{lfs} module, we could not type two overloaded functions
due to the lack of intersection types in our type system.

@@

However, we hit some limitations of our type system while annotating this module.
First, it uses the function \texttt{getmetatable} to get a table and
checks whether this table has the field \texttt{\string_\string_type} or not.
We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
type \texttt{any} as its return type, and this generates a warning.
Second, the module returns a metatable that extends \texttt{\string_\string_call}
with \texttt{typical.type}, that is, we can use the module itself as a function,
though it is a table.
Our type system still does not support metatables, so we did not extend our version
of the \texttt{typical} module to support \texttt{\string_\string_call}.
Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
We could not precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In this module, we could only type the two error reporting functions
that it exports.

@@

Even though we could precisely type the interface that most of
the modules export, we had issues to write mutually recursive functions.
This kind of functions often appear on compilers construction
to traverse the data structures that they use.
However, Typed Lua still does not support mutually recursive functions.
A way to overcome this limitation was to predeclare these functions
with an empty body, and then redeclare them with they actual body.
The first declaration specifies the function type, while the
second specifies what the function actually does without changing
any type definition.

>>>>

LuaSocket \citep{luasocket} and LuaFileSystem \citep{luafilesystem} were
the third and fourth case studies that we used to evaluate Typed Lua.
We chose them because they are the most popular Lua libraries.
We wrote a script that builds the dependency graph of Lua libraries
that are in the LuaRocks repository, and uses this dependency graph
to identify the most popular Lua libraries.

We also randomly selected three case studies from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first provides client side HTTP digest authentication for Lua.
The second is an extension to the primitive function \texttt{type}.
The third is a generator and checker of modulo 11 numbers.
We randomly selected three case studies because we wanted to evaluate
Typed Lua for annotating existing libraries that are written in Lua,
as the previous case studies are mostly libraries that are written in C.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
Lua Standard Library & 62\% & 6\% & 9\% & 23\% & 129 \\ % 100%
\hline
MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
\hline
LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them include.
We use these split results to discuss the contributions and
limitations of our type system.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
base & 35\% & 12\% & 8\% & 45\% & 26 \\ % 100%
\hline
coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
\hline
package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
\hline
string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
\hline
utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
\hline
table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
\hline
math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
\hline
io & 71\% & 0\% & 10\% & 19\% & 21 \\ % 100%
\hline
os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

The \texttt{base} module was very difficult to type
because it includes several functions that rely on reflection,
as the \emph{hard} category shows.
For instance, the functions \texttt{pairs} and \texttt{getmetatable}
are in this category.
While \texttt{pairs} traverses all keys and values that are stored
in a given table, \texttt{getmetatable} returns the metatable of a
given table.

There are some functions in the \texttt{base} module that
we could not give a precise static type because our type
system does not have parametric polymorphism,
as the \emph{poly} category shows.
This is the case of \texttt{ipairs}, \texttt{rawget}, and
\texttt{rawset}.

The \texttt{base} module also includes some overloaded functions,
as the \emph{over} category shows.
We could not type these functions because our type system does
not include intersection types.
This is the case of \texttt{tonumber} and \texttt{collectgarbage}.

@@

The \texttt{coroutine} module was also very difficult to type
because our type system cannot describe the computational effects
of a program.
The \emph{hard} category shows the amount of functions that we
could not give a precise static type for this reason.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.

Still, we could give a precise static type to one function in
the \texttt{coroutine} module, as the \emph{easy} category shows.
The function \texttt{coroutine.isyieldable} has no input parameters,
and it simply returns a boolean that indicates whether the running
coroutine is yieldable.

We could give precise static types to the constants and functions
of the \texttt{package} module, but we could not give precise static
to the following tables:
\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types rely on reflection,
that is, their types depend on the modules a program loads.
For this reason, they are in the \emph{hard} category.

We could give precise static types to most of the functions of the
\texttt{string} module, but we could not give precise static types
to the functions that rely on format strings.
For instance, the type of the arguments that we pass to
\texttt{string.format} must match the format string we are using.
It is fine to call \texttt{string.format("\%d", 1)}, but
\texttt{string.format("\%d", true)} raises a run-time error.
These functions that rely on format strings are in the \emph{hard} category.

Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style.
For instance, \texttt{string.len("foo")} and \texttt{("foo"):len()}
produce the same result.
We did not include the typing of the \texttt{string} methods in our
evaluation results, because we treat this feature as a special case
in the implementation.

@@

The \texttt{table} module was specially difficult to type because
most of its functions require parametric polymorphism,
as the \emph{poly} category shows.
These functions either receive or return a list of elements, and
parametric polymorphism would help us to describe them with a generic type.

However, the lack of parametric polymorphism did not prevent us from
giving a precise type to one function of the \texttt{table} module,
as the \emph{easy} category shows.
We could give a precise static type to \texttt{table.concat},
as it operates over lists where all elements are strings or numbers.

Even if our type system had parametric polymorphism, there is
still one function of the \texttt{table} module that we could
not give a precise static type because it is an overloaded function,
as the \emph{over} category shows.
This function is \texttt{table.insert}, and its type depends on
the calling arity.
More precisely, calling \texttt{table.insert(l, v)} inserts the
element \texttt{v} at the end of the list \texttt{l},
while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
at the position \texttt{p} of the list \texttt{l},
but generates a run-time error when \texttt{p} is out of bounds.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

Even though the \texttt{math} module looks straightforward to type,
the \emph{over} category shows that it includes several overloaded functions.
For instance, the functions \texttt{math.abs} and \texttt{math.random}
are in this category.

In the case of \texttt{math.abs},
calling \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.
Even though we gave this function the type \texttt{(number) -> (number)},
it is not precise enough because the return type is always \texttt{number}
independently of the argument type.
In other words, the return type should be \texttt{integer} when
the argument type is also \texttt{integer}, and the return type
should be \texttt{number} when the argument type is also \texttt{number}.

In the case of \texttt{math.random}, its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
Like \texttt{table.insert}, this function also does not follow the
semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

The \texttt{io} module provides operations for manipulating files,
and these operations can use implicit or explicit file descriptors.
The implicit operations are functions in the \texttt{io} table,
while the explicit operations are methods of a file descriptor.
We used an \emph{userdata} declaration to introduce the type
\texttt{file} for representing the type of a file descriptor
and its methods.
The evaluation results include both implicit and explicit operations.

We could give precise static types to most of the members of the
\texttt{io} module, but its categories \emph{over} and \emph{hard}
show that it includes several members that we could not give a precise
static type.
While the function \texttt{io.close} and the method \texttt{file:close}
are in the \emph{over} category,
the functions \texttt{io.read} and \texttt{io.lines} are
in the \emph{hard} category along with the methods
\texttt{file:read} and \texttt{file:lines}.

We could not precisely type \texttt{io.close} because it is
an overloaded function.
This means that its return type depends on whether the file handle
that is being closed was created with \texttt{io.popen}.
Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
when \texttt{f} was created with \texttt{io.popen},
but it returns the tuple \texttt{(boolean)?} otherwise.
The method \texttt{file:close} has the same issue.

We could not precisely type \texttt{io.read} because its return type
relies on format strings.
For instance, calling \texttt{io.read("l")} returns a string or \texttt{nil},
\texttt{io.read("n")} returns a number or \texttt{nil}, and
\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
The function \texttt{io.lines}, and the methods \texttt{file:read} and
\texttt{file:lines} have the same issue.

There are two functions in the \texttt{os} module that we could
not give a precise static type because they are overloaded functions.
The functions that are in the \emph{over} category are
\texttt{os.date} and \texttt{os.execute}.

@@

We could give precise static types to most of the members
in the \texttt{socket} module, which is the C core.
However, this module includes some functions that
we could not give a precise static type because they
rely on reflection, as the \emph{hard} category shows.
For instance, \texttt{socket.skip} is a function that is
in this category.
We can use this function to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} does not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \citep{flanagan2006htc}.

We could give precise static types to most of the members of the
modules \texttt{ftp} and \texttt{http}, but we could not precisely type
two overloaded functions: \texttt{ftp.get} and \texttt{http.request}.

@@

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encoding.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low-level
filters to create standard filters.

@@

Even though we could precisely type most of the functions exported
by the \texttt{lfs} module, we could not type two overloaded functions
due to the lack of intersection types in our type system.

@@

However, we hit some limitations of our type system while annotating this module.

First, it uses the \texttt{getmetatable} to get a table and
checks whether this table has the field \texttt{\string_\string_type}.
We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
type \texttt{any} as its return type, and this generates a warning.

Second, it returns a metatable that extends \texttt{\string_\string_call}
with \texttt{typical.type}, that is, we can use the module itself as a function,
though it is a table.
Our type system still does not support metatables, so we did not extend our version
of the \texttt{typical} module to support \texttt{\string_\string_call}.

Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
As we mentioned in this chapter, we use the dynamic type as a replacement for
the lack of type parameters.
This means that we get warnings inside an \texttt{ipairs} iteration because
all iterated elements have the dynamic type.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
Even though LPeg is the third most popular Lua module,
we cannot not precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In the \texttt{tllexer} module, we could only give precise static types
to two error reporting functions that it exports.

@@

Even though we could precisely type the interface that most of
the modules export, we had issues to write mutually recursive functions.
This kind of functions often appear in compilers construction
to traverse the data structures that they use.
However, Typed Lua still does not support mutually recursive functions.
A way to overcome this limitation was to predeclare these functions
with an empty body, and then redeclare them with they actual body.
The first declaration specifies the function type, while the
second specifies what the function actually does without changing
any type definition.

### 2015-02-23 doc/thesis/related.tex

In this chapter we review related work as a way to indentify the novel
features of our type system, what are the features that it shares with
other type systems, and whether we can incorporate some interesting
features from other type systems.
We split this chapter into two parts: in the first part we review
other Lua projects, while in the second part we review other
projects that are not related to Lua.

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
Tidal Lock covers a little subset of Lua.
Statements include declaration of local variables, multiple assignments,
function application, and the return statement.
This means that Tidal Lock does not include any control-flow statement.
Expressions include primitive literals, table indexing, function application,
function declaration, and the table constructor, but they do not include
binary operations.

@@

@@

Ravi \citep{ravi} is an experimental Lua dialect.
Ravi introduces optional static typing for Lua to improve run-time performance.
To do that, Ravi extends the Lua Virtual Machine to include new
operations that take into account static type information.

@@

>>>>

In this chapter we review related work as a way to identify the novel
features of our type system, what are the features that it shares with
other type systems, and whether we can incorporate some interesting
features from other type systems.
We split this chapter into two sections: in the first section we review
other Lua projects, while in the second section we review other
projects that are not related to Lua.

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
Tidal Lock covers a little subset of Lua.
Statements include declaration of local variables, multiple assignment,
function application, and the return statement.
This means that Tidal Lock does not include any control-flow statement.
Expressions include primitive literals, table indexing, function application,
function declaration, and the table constructor, but they do not include
binary operations.

@@

Sol \citep{sol} is an experimental optional type system for Lua.
Its type system is similar to ours, as it includes literal types,
union types, and function types that handle variadic functions.
However, it does not handle the refinement of tables and it
includes different types for tables.
Sol types tables as lists, maps, and objects.
Its object types handle a specific object-oriented idiom that
Sol introduces.

Lua Analyzer \citep{luaanalyzer} is an optional type system for Lua
that is specially designed to work in the Löve Studio,
an IDE for game developing using the Löve framework.
It works in Lua 5.1 only, and uses type annotations inside comments.
It is unsound by design because its dynamic type is both
top and bottom in the subtyping relation.

Lua Analyzer shares some features with Typed Lua, and also
has some interesting features that we do not have in Typed Lua.
It has similar rules for handling the \texttt{or} idiom and
discriminating union types inside conditions.
However, these rules are limited to the \texttt{nil} tag only.
It also includes different types for typing tables.
It includes regular record types that maps names to types,
array types, and map types.
Even though it does not support the refinement of tables,
it allows the definition of nominal table types that simulate classes.
This system allows it to type check custom class systems,
which are common in Lua.
Function types also support multiple return values and
variadic functions, but they do not support overloading the
return type.
Recently, it included experimental support for type aliases and generics.

@@

Ravi \citep{ravi} is an experimental Lua dialect.
Ravi introduces optional static typing for Lua to improve run-time performance.
To do that, Ravi extends the Lua Virtual Machine to include new
operations that take into account static type information.
Currently, Ravi extends the Lua Virtual Machine to support few types:
\texttt{int}, \texttt{double}, arrays of integers, and arrays of doubles.

@@

Mypy \citep{mypy} is an optional type system for Python.
The type system of mypy is similar to the type system of
Reticulated Python, but mypy does not insert run-time checks
and it has parametric polymorphism.
In contrast, Reticulated Python can type variadic functions,
but mypy cannot.
Recently, Guido van Rossum, Python's author, proposed a
standard syntax for type annotations in Python \citep{PEP483}
that is extremely inspired by mypy \citep{PEP484}.
The main goal of this proposal is to make easier building
static analysis tools for Python.
Typing \citep{typing} is a tool that is being developed to
implement this proposal.
